Species motif participation provides unique information about species risk of extinction

### 2022-06-07 manuscript/response_PRSB_round2.tex

        \them
        Line 289: Odd use of direct competition when most other papers use exploitative competition

        \us We are not aware of any other motif-focused papers calling this motif ``exploitative competition''. Our terminology follows the nomenclature from the introduction of motifs to food webs~\citep{Stouffer2007}.
        Note that ``direct competition'' refers simply to two species which consume the same prey and is agnostic as to whether this is exploitative or interference competition. 
        As exploitative and interference competition are not the same ecologically, we have maintained the conventional motif terminology so as not to imply an overly specific interpretation.

        \them
        Line 334: in-degree and trophic level are not sufficient to explain the trends, yet there is a strong negative correlation between apparent competition and in-degree. This seems contradictory.

        \us The Reviewer is correct that there is a strong  correlation between apparent competition and in-degree, as we state in the text. It is also true that apparent competition and in-degree show opposite relationships with persistence, such that the apparent competition-persistence relationship could be described by the in-degree-persistence relationship. However, as detailed later in the paragraph, this ignores the strong correlation between apparent competition and trophic level which would suggest that persistence should show different relationships to apparent competition. This means that either trophic level is unrelated to persistence (which we know is not true) or some other factor is over-riding the ``effect" (there is no clear causality from trophic level to motifs or vice versa) of trophic level on apparent competition.

        We have revised this paragraph to stop readers from jumping to conclusions before reading the entire section.

        \begin{quotation}
            For apparent competition, there were strong and opposing correlations with in-degree and trophic level but the relationship between apparent competition and persistence only matches the trends for in-degree.
            This suggests that some other factor is over-riding the relationship between apparent competition and trophic level.        
        \end{quotation}

### 2022-06-07 manuscript/reorder_main_text.tex
trying to fix R1s poor reading comprehension
Updates from Overleaf

    %\subsection*{2 Global and local network properties are known to affect stability}
    How disturbances, such as population decline or extinction, lead to loss of other species via direct and indirect effects (i.e., secondary extinctions) is a vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions \citep{Dunne2002, Eklof2006} because consumers, on average, have access to a larger number of prey species. However, in a well-connected network a disturbance may spread more easily and cause more secondary extinctions due to the presence of many possible pathways \citep{}.
    The net effect of connectance on extinction therefore likely depends on the severity of disturbance and the ability of consumers to subsist on a reduced set of prey.

    Local properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, including secondary extinction due to reduced plant populations.

    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    As well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may, for example, be at greater risk of extinction than a specialist which consumes a highly generalist resource that is very unlikely to go extinct.

@@

    While the potential relationships between motif participation and species-level extinction risk have not yet been tested to our knowledge, it has been shown that the frequencies of different motifs within a whole network are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}. 
    Given these relationships between network-level motif frequencies and overall community stability, it is worth investigating whether a species' participation in different motifs also affects its extinction risk.
    We are also interested in whether the relationships between network structure and extinction risk depend on the severity of disturbance experienced by the community.

    Naturally, a larger disturbance tends to lead to more secondary extinctions \citep{}.
    Disturbances to plant species can also affect some consumers more strongly than others  \citep{byrnes2011climate}. 
    In empirical studies of kelp forests, the magnitude of the disturbances to basal resources can alter the competition between species in the benthic community \citep{detmer2021variation} and change the percent coverage of different taxonomic groups \citep{carnell2020more}.

    %While these studies uncovered the potential influence of network-level motif frequencies on network stability, a disturbance may likely alter the network and the interactions between species in different ways. 
    %This can affect the influence of motif frequencies on network stability, as it has been demonstrated that omnivory may appear stabilizing dependent on interactions with other species \citep{Kratina2012}.

%\subsection*{Here we test...}
   Here, we are interested in whether a consumer's extinction risk of going secondarily extinct depends on 1) the motifs in which the species participates, 2) the motif profile of the full food web and 3) the severity of the disturbance to the basal level. %We address this in scenarios with various levels of disturbance targeting the basal level.
    In addition, we want to understand how the information provided by motifs is related to global and local food web properties with known relationships to extinction risk in order to better synthesize information provided by different scales of network structure.
    We model secondary extinction probabilities using Bayesian networks \citep{Eklof2013, Haussler2020}. 
    We focus our analyse on the four network motifs that are over-represented in empirical networks and additionally have been shown to contribute to network stability; the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.

     %\subsection*{2 Bayesian networks are a fantastic way to model bottom-up effects}

@@

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function "nichemodel" within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw}. 
		To capture a range of plausible network architectures, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details, se \emph{Appendix S1}.
		We removed any biologically unlikely networks (Appendix \emph{S1}) and replaced them with new simulated networks, repeating the process until we obtained 100 suitable networks in each combination of size and connectance.

@@

        As consumers do not affect the extinction probabilities of their resources, Bayesian networks can only capture bottom-up effects of disturbances within networks. 
        However, Bayesian networks do capture the majority of secondary extinctions captured by dynamical models \citep{Eklof2013}.
        This implies that bottom-up effects drive most secondary extinctions in simulations, echoing the strong bottom-up effects observed in empirical studies~\citep{}.
        [[Something about strong bottom-up effects in empirical studies]]
    	To make the simulated networks compatible with the Bayesian simulation framework, all networks were rendered acyclic following~\citet{Allesina2009functional} and topologically sorted from lowest to highest trophic level (Appendix \emph{S1}).
    	These processing steps allow the strictly bottom-up calculation of probability of persistence.

@@

        We are primarily interested in whether a consumer's extinction risk is related to its motif participation or the motif profile of the network and, secondarily, also wish to connect the information provided by motifs to global and local network properties that have known effects on extinction risk.
    	To address these questions, we tested 1) how species persistence varied with motif participation, 2) how motif participation varied with other network properties (in-degree and STL), 3) how network mean persistence (the average persistence of consumers) varied with network motif profiles, 4) how network motif profiles varied with global structural properties (S and C), and 5) how species persistence and network mean persistence varied with other network properties (in-degree, STL, S, C, and level of disturbance).
    	See Appendix \emph{3.1 - 3.5} for details. 
    	All analyses are briefly described below and were performed using the R~\citep{R} packages \emph{lmerTest}~\citep{lmerTest}, \emph{MuMIn}~\citep{MuMIn}, and \emph{vegan}~\citep{vegan}.

@@

            Our main question of interest is whether appearing in different motifs is associated with differences in extinciton risk. 
            To test if participating more often in a certain motif affected a species' probability of persistence, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks):
                \begin{equation}
                    \Psi_{ink} \approx \rho_{i} + \pi_{k} + \rho_{i}\pi_{k} +
                    S_{n}C_{n} + N_n,
                    \label{propreq}
                \end{equation}
            \noindent where $\Psi_{ink}$ is the persistence of species $i$, belonging to network $n$, during disturbance level $k$, $\rho_{i}$ is the proportion of the role of species $i$ that is made up by the focal motif, $\pi_k$ is the probability of extinction for a basal resource in disturbance level $k$,  $S_{n}C_{n}$ is a random intercept for the species richness and connectance of network $n$, and $N_n$ is a random effect of network $n$, intended to account for the non-independence of motif participations for species in the same network.
            All predictors were centered and scaled before fitting the models.         
            To test whether these general trends were consistent across networks, we visually examined the distribution of slopes of simplified regressions fit to a single network and level of disturbance (Appendix \emph{3.1}).

@@

            Properties which refer to different scales of network structure (e.g., local, global, and meso-scale) are not independent. 
            This means that, to understand relationships between motif participation (a meso-scale network property) and extinction risk, we must also consider how motif participation varies with other properties known to affect extinction risk.
            To identify correlations between global network structure (i.e., the whole network) and participation in particular motifs, we fit a series of LMs relating motif participation to network size, connectance, and their interaction (Appendix \emph{S3.2}).
            We fit a similar set of  LMMs relating motif participation to in-degree or trophic level (i.e., local measures of a species' role; Appendix \emph{S3.2}).
            Taken together, these supplemental models demonstrate how motif participation relates to other measures of network structure.

@@

            As well as relating species-specific probabilities of extinction to species-specific motif participation, we are interested in whether the overall frequency of motifs in a network provides information about the overall probability of any species going extinct.
            Just as other whole-network properties (S, C, etc.) are known to affect the overall stability of a network, it is possible that the arrangement of interactions into different motifs could reflect different average probabilities of persistence across all consumers.
            We tested whether networks with similar motif profiles tend to have similar mean persistences among consumers using a PERMANOVA (Appendix \emph{S3.3}).
            To identify the relationships between particular motifs and network mean persistence in more detail, we fit a set of four LMs (one per motif):
                \begin{equation}
                    \overline{\Psi_{nk}} \approx \Bar{\rho}_{n} + \pi_{k} + \Bar{\rho}_{n}\pi_{k},
                    \label{netpropeq}
                \end{equation}
            \noindent where $\overline{\Psi_{nk}}$ is the mean persistence of all consumers in network $n$ at disturbance level $k$, $\Bar{\rho}_{n}$ is the proportion of the focal motif in the network's motif profile, and $\pi_k$ is the level of disturbance (probability of extinction for basal resources) as in equation~\ref{propreq}.

@@

            Just like motif participation, a network's motif profile is not likely to be independent of other levels of structure. 
            To formally test whether networks with the the same size and connectance have more similar motif profiles, we used a PERMANOVA~\citep{Anderson2001} of Bray-Curtis dissimilarity among motif profiles against network size, connectance, and their interaction (Appendix \emph{S3.4}). 
            In addition, to understand how motif profiles may vary with global network structure in more detail, we identify correlations between size, connectance, and particular motifs using a set of four LMs:
            \begin{equation}
                \Bar{\rho_{n}} \approx \Sigma_{n} + \zeta_{n} + \Sigma_{n}\zeta_{n}
            \label{profile_global}
            \end{equation}
            \noindent where $\Bar{\rho}_{n}$ is the proportion of the focal motif in the motif profile of network $n$ and $\Sigma_n$ and $\zeta_{n}$ are the size and connectance of network $n$.

>>>>

    %\subsection*{2 Global and local network properties are known to affect stability}
    How disturbances, such as population decline or extinction, lead to loss of other species via direct and indirect effects (i.e., secondary extinctions) is a vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions because consumers, on average, have access to a larger number of prey species and can switch from a declining or extinct prey to another resource~\citep{Dunne2002, Eklof2006,Baumgartner2020}. However, a well-connected network also includes more pathways through which a disturbance can spread~\citep{Dunne2004,Vieira2015}.
    Depending on the strength and arrangements of interactions, this means that species may be impacted through several paths, amplifying the original disturbance and causing more secondary extinctions.
    The net effect of connectance on extinction therefore likely depends on finer-scale network structure, as well as the type and severity of disturbance.

    Local properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, including secondary extinction due to reduced plant populations.
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    As well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may, for example, be at greater risk of extinction than a specialist which consumes a highly generalist resource that is very unlikely to go extinct.

@@

    While the potential relationships between motif participation and species-level extinction risk have not yet been tested to our knowledge, it has been shown that the frequencies of different motifs within a whole network are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}. 
    These relationships between network-level motif frequencies (`motif profiles') and overall community stability mean that a network's motif profile could affect the average extinction risk for consumers within the network.
    Moreover, if network motif profiles are related to average extinction risk, it is worth investigating whether a species' participation in different motifs also affects its extinction risk.

    As mentioned above, the severity of a disturbance may interact with network structure to affect secondary extinctions.
    A mild disturbance (e.g., a small increase in the probability of basal resources going extinct) is intuitively likely to cause fewer secondary extinctions than a severe disturbance, and the difference in effects can depend on network structure~\citep{Baumgartner2015}.
    Empirical studies have shown that disturbances to plant species can also affect some consumers more strongly than others  \citep{byrnes2011climate}, and that the strength of disturbance also affects which consumer species are most strongly affected~\citep{detmer2021variation,carnell2020more}. 
    By taking advantage of recent advances in rapid simulation of ecological networks, we are able to simulate a range of disturbance intensities and test whether meso-scale network structure also interacts with disturbance.

%\subsection*{Here we test...}
    Here, we are interested in whether a consumer's extinction risk of going secondarily extinct depends on 1) the motifs in which the species participates, 2) the motif profile of the full food web and 3) the severity of the disturbance to the basal level. %We address this in scenarios with various levels of disturbance targeting the basal level.
    In addition, we want to understand how the information provided by motifs is related to other global and local food web properties in order to better synthesize information provided by different scales of network structure.
    We model secondary extinction probabilities using Bayesian networks \citep{Eklof2013, Haussler2020}. 
    We focus our analyse on the four network motifs that are over-represented in empirical networks and additionally have been shown to contribute to network stability; the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.

@@

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function "nichemodel" within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw}. 
		To capture a range of plausible network architectures, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details, see \emph{Appendix S1}.
		We removed any biologically unlikely networks (Appendix \emph{S1}) and replaced them with new simulated networks, repeating the process until we obtained 100 suitable networks in each combination of size and connectance.

@@

        As consumers do not affect the extinction probabilities of their resources, Bayesian networks can only capture bottom-up effects of disturbances within networks. 
        However, Bayesian networks do capture the majority of secondary extinctions captured by dynamical models \citep{Eklof2013}.
        [[Something about strong bottom-up effects in empirical studies]]
    	To make the simulated networks compatible with the Bayesian simulation framework, all networks were rendered acyclic following~\citet{Allesina2009functional} and topologically sorted from lowest to highest trophic level (Appendix \emph{S1}).
    	These processing steps allow the strictly bottom-up calculation of probability of persistence.

@@

    	We tested 1) how species persistence varied with motif participation, 2) how motif participation varied with other network properties (in-degree and STL), 3) how network mean persistence (the average persistence of consumers) varied with network motif profiles, 4) how network motif profiles varied with global structural properties (S and C), and 5) how species persistence and network mean persistence varied with other network properties (S, C, in-degree, and trophic level).
        Where applicable, we also consider how interactions between network structure and disturbance level might affect extinction risk.
    	See Appendix \emph{3.1 - 3.5} for details. 
    	All analyses are briefly described below and were performed using the R~\citep{R} packages \emph{lmerTest}~\citep{lmerTest}, \emph{MuMIn}~\citep{MuMIn}, and \emph{vegan}~\citep{vegan}.

@@

            To test if participating more often in a certain motif affected a species' probability of persistence, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks):
                \begin{equation}
                    \Psi_{ink} \approx \rho_{i} + \pi_{k} + \rho_{i}\pi_{k} +
                    S_{n}C_{n} + N_n,
                    \label{propreq}
                \end{equation}
            \noindent where $\Psi_{ink}$ is the persistence of species $i$, belonging to network $n$, during disturbance level $k$, $\rho_{i}$ is the proportion of the role of species $i$ that is made up by the focal motif, $\pi_k$ is the probability of extinction for a basal resource in disturbance level $k$,  $S_{n}C_{n}$ is a random intercept for the species richness and connectance of network $n$, and $N_n$ is a random effect of network $n$, intended to account for the non-independence of motif participations for species in the same network.
            All predictors were centered and scaled before fitting the models.         
            To test whether these general trends were consistent across networks, we visually examined the distribution of slopes of simplified regressions fit to a single network and level of disturbance (Appendix \emph{3.1}).

@@

            Properties which refer to different scales of network structure, including different amounts of direct and indirect interactions with the focal species, are not independent. 
            To identify correlations between global network structure (i.e., the whole network) and participation in particular motifs (a meso-scale description of species roles), we fit a series of LMs relating motif participation to network size, connectance, and their interaction (Appendix \emph{S3.2}).
            We fit a similar set of  LMMs relating motif participation to in-degree or trophic level (i.e., local measures of a species' role; Appendix \emph{S3.2}).
            Taken together, these supplemental models demonstrate how motif participation relates to other measures of network structure.

@@

            As well as species-specific probabilities of persistence, the overall frequency of motifs in a network may provide information about the overall probability of any species going extinct (since they incorporate information on indirect effects).
            We tested whether networks with similar motif profiles tend to have similar mean persistences among consumers using a PERMANOVA (Appendix \emph{S3.3}).
            To identify the relationships between particular motifs and network mean persistence in more detail, we fit a set of four LMs (one per motif):
                \begin{equation}
                    \overline{\Psi_{nk}} \approx \Bar{\rho}_{n} + \pi_{k} + \Bar{\rho}_{n}\pi_{k},
                    \label{netpropeq}
                \end{equation}
            \noindent where $\overline{\Psi_{nk}}$ is the mean persistence of all consumers in network $n$ at disturbance level $k$, $\Bar{\rho}_{n}$ is the proportion of the focal motif in the network's motif profile, and $\pi_k$ is the level of disturbance (probability of extinction for basal resources) as in equation~\ref{propreq}.

@@

            To test whether networks with the the same size and connectance have more similar motif profiles, we used a PERMANOVA~\citep{Anderson2001} of Bray-Curtis dissimilarity among motif profiles against network size, connectance, and their interaction (Appendix \emph{S3.4}). 
            In addition, we identify correlations between size, connectance, and particular motifs using a set of four LMs:
            \begin{equation}
                \Bar{\rho_{n}} \approx \Sigma_{n} + \zeta_{n} + \Sigma_{n}\zeta_{n}
            \label{profile_global}
            \end{equation}
            \noindent where $\Bar{\rho}_{n}$ is the proportion of the focal motif in the motif profile of network $n$ and $\Sigma_n$ and $\zeta_{n}$ are the size and connectance of network $n$.
